---
title: "gdp_hspend"
author: "Duzhi Zhao"
date: "11/19/2019"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    code_folding: hide
editor_options: 
  chunk_output_type: console
---
# Dataset: MMR vs GDP PER CAPITA
```{r setup, include=FALSE}
library(tidyverse)
library(viridis)
library(readxl)
library(plotly)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8, 
  fig.height = 6,
  out.width = "100%"
)
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

## Import dataset
```{r}
# Maternal Mortality Ratio: number of maternal deaths during a given time period per 100,000 live births during the same time period.
# MMR measures the obsteric risk a women is facing when she is pregnant. 
# E.g. 1575 per 100,000 live births (Afghanistan) vs. 5 per 100,000 live births (Sweden) means a women in Afghanistan is facing a 315 times higher risk of maternal death when she is pregnant than a woman in Swedan. 
mmr_gdp_capita = read_csv("./data/maternal-mortality-ratio-vs-gdp-per-capita.csv") %>%
  janitor::clean_names() %>% 
  rename(
    country = entity,
    mmr = maternal_mortality_ratio_deaths_per_100_000_live_births,
    gdp_per_capita = gdp_per_capita_international,
    gdp = x6) %>% 
  mutate(
    country = as.factor(country),
    code = as.factor(code)
  ) %>% 
  mutate(
    gdp_per_capita = round(gdp_per_capita, digits = 2)
  ) %>% 
  filter(year > 1999, year < 2016)

summary(is.na(mmr_gdp_capita$mmr)) # 945 NA, 3435 non-NA
summary(is.na(mmr_gdp_capita$gdp_per_capita)) # 845 NA, 3535 non-NA
summary(is.na(mmr_gdp_capita$gdp)) # 750 NA, 3630 non-NA
```

## log-transformed 'gdp_per_capita'
```{r}
# log-transformed
mmr_gdp_capita_log = 
  mmr_gdp_capita %>% 
  mutate(
    ln_gdp_per_capita = round(log(gdp_per_capita), digits = 2)
  )
```

# Maternal mortality ratio vs. GDP per capita
```{r}
x1_template = list(
  title = "GDP per capita ($)"
)

x2_template = list(
  title = "ln(GDP per capita)"
)

y_template = list(
  title = "Maternal Mortality Ratio per 100,000 live births"
)

mmr_gdp_capita_log %>%  
  filter(year == 2013) %>% 
  plot_ly(
    x = ~gdp_per_capita, y = ~mmr, type = "scatter", 
    mode = "markers", color = ~country
  ) %>%  
  layout(
    xaxis = x1_template,
    yaxis = y_template,
    title = "Maternal mortality ratio vs. GDP per capita"
  )
```

# Maternal mortality ratio vs. ln(GDP per capita)
```{r}
mmr_gdp_capita_log %>%  
  filter(year == 2013) %>% 
  plot_ly(
    x = ~ln_gdp_per_capita, y = ~mmr, type = "scatter", 
    mode = "markers", color = ~country
  ) %>%  
  layout(
    xaxis = x2_template,
    yaxis = y_template,
    title = "Maternal mortality ratio vs. ln(GDP per capita)"
  )
```

# Maternal mortality ratio in top 10 developed countries
```{r}
ten_developed_countries_gdp_per_capita = 
mmr_gdp_capita_log %>%  
  filter(year == 2013,
         country %in% c("United States", 
                        "Switzerland", 
                        "Germany",
                        "Singapore",
                        "Japan",
                        "Canada",
                        "United Kingdom",
                        "Spain",
                        "France", 
                        "Australia"
                        )) %>% 
  mutate(country = as.factor(country)) %>% 
  ggplot(aes(x = country, y = gdp_per_capita, fill = country)) +
  geom_col(position = "dodge") +
  coord_flip() +
  theme(legend.position = "none") +
  labs(x = "Developed countries",
       y = "GDP per capita (dollars)",
       title = "Cost of child birth in 10 developed countries")

interactive_gdp_per_capita = ggplotly(ten_developed_countries_gdp_per_capita, 
                                          tooltip = c("x", "y"))
```

# Healthcare spending (%GDP) vs. Country

## Import and clean data
```{r}
# Percent GDP that comes from health spending
health_spending_percent_gdp = read_csv("./data/WorldBank/health_spending_percent_GDP.csv") %>%
  janitor::clean_names() %>% 
  select(country_name, country_code, x2000:x2015) %>% 
  pivot_longer(
  x2000:x2015,
  names_to = "year",
  values_to = "health_spending_percent_GDP") %>% 
  rename("health_spending_percent_gdp" = "health_spending_percent_GDP") %>% 
  mutate(year = str_remove(year,"x")) %>% 
  mutate(health_spending_percent_gdp = round(health_spending_percent_gdp, digits = 2)) %>% 
  mutate(country_name = as.factor(country_name),
         country_code = as.factor(country_code),
         year = as.numeric(year)) %>% 
  rename(country = country_name,
         code = country_code)
```

## Filtered dataset to only include 10 developed countries in 2013
```{r}
healthcare_spending_2013_developed = 
  health_spending_percent_gdp %>% 
  filter(year == 2013) %>% 
  filter(country %in% c("United States", 
                        "Switzerland", 
                        "Germany",
                        "Singapore",
                        "Japan",
                        "Canada",
                        "United Kingdom",
                        "Spain",
                        "France", 
                        "Australia"
                        ))
```

## Bar chart (% GDP that comes from healthcare spending in 10 developed countries)
```{r}
ten_developed_countries_health_spending = 
  healthcare_spending_2013_developed %>% 
  mutate(country = as.factor(country),
         country = fct_reorder(country, health_spending_percent_gdp)) %>% 
  ggplot(aes(x = country, y = health_spending_percent_gdp, fill = country)) +
  geom_col(position = "dodge") +
  coord_flip() +
  theme(legend.position = "none") +
  labs(x = "Developed countries",
       y = "Percent GDP that comes from healthcare spending(%)",
       title = "Different amount of healthcare spending across 10 developed countries")

interactive_healthcare_spending = ggplotly(ten_developed_countries_health_spending,
                                           tooltip = c("y", "x", "colour"))
```

# Maternal mortality across 10 developed countries

## Import and clean data
```{r}
# Maternal Mortality ratio 
maternal_mort = read_csv("./data/Gapminder/maternal_mortality_ratio_per_100000_live_births.csv") %>% 
  janitor::clean_names() %>% 
  pivot_longer(
    x1800:x2013,
    names_to = "year",
    values_to = "percent_maternal_mortality"
  ) %>% 
  mutate(year = str_remove(year,"x"))
```

## Filtered dataset to only include 10 developed countries in 2013
```{r}
maternal_mort_2013_developed = 
  maternal_mort %>% 
  filter(year == 2013) %>% 
  filter(country %in% c("United States", 
                        "Switzerland", 
                        "Germany",
                        "Singapore",
                        "Japan",
                        "Canada",
                        "United Kingdom",
                        "Spain",
                        "France", 
                        "Australia"
                        ))
```

## Bar chart (maternal mortality in 10 developed countries)
```{r}
ten_developed_countries_maternal_mort = 
  maternal_mort_2013_developed %>% 
  mutate(country = as.factor(country),
         country = fct_relevel(country, 
                               "Singapore",
                               "Australia",
                               "Spain",
                               "United Kingdom",
                               "Canada",
                               "Japan",
                               "Germany",
                               "Switzerland",
                               "France",
                               "United States")) %>% 
  ggplot(aes(x = country, y = percent_maternal_mortality, fill = country)) +
  geom_col(position = "dodge") +
  coord_flip() +
  theme(legend.position = "none") +
  labs(x = "Developed countries",
       y = "Maternal mortality(%)",
       title = "Mismatching between US healthcare spending and maternal mortality")

interactive_maternal_mortality = ggplotly(ten_developed_countries_maternal_mort, 
                                          tooltip = c("x", "y"))
```


# Cost of giving birth in 10 developed countries

## Import and clean data
```{r}
# cost of giving birth
child_birth_cost = read_csv("./data/cost_of_child_birth_global.csv") %>% 
  janitor::clean_names() %>% 
  mutate(average_cost = (child_birth_cost_normal + child_birth_cost_csec)/2)
```

## Bar chart (cost of childbirth in 10 developed countires)
```{r}
ten_developed_countries_childbirth_cost = 
  child_birth_cost %>% 
  mutate(country = as.factor(country)) %>% 
  ggplot(aes(x = country, y = average_cost, fill = country)) +
  geom_col(position = "dodge") +
  coord_flip() +
  theme(legend.position = "none") +
  labs(x = "Developed countries",
       y = "Average cost of giving birth(dollars)",
       title = "Cost of child birth in 10 developed countries")

interactive_child_birth_cost = ggplotly(ten_developed_countries_childbirth_cost, 
                                          tooltip = c("x", "y"))
```


## Combine three interactive graphs
```{r}
subplot(interactive_healthcare_spending, 
        interactive_maternal_mortality,
        shareX = T, shareY = T) %>% 
  layout(title = "Mismatch between %GDP that comes from healthcare spending and maternal mortality ratio")


subplot(interactive_child_birth_cost,
        interactive_gdp_per_capita, 
        shareX = T, shareY = T) %>% 
  layout(title = "Top 10 developed countries")
  
```




# Health staff
```{r}
#births attended by skilled health staff (percent of total)
staff = read_csv("./data/Gapminder/births_attended_by_skilled_health_staff_percent_of_total.csv")  %>% 
  janitor::clean_names() %>% 
  select(country, x2000:x2017) %>% 
  pivot_longer(
    x2000:x2017,
    names_to = "year",
    values_to = "perc_births_attended"
  ) %>% 
  mutate(year = str_remove(year,"x")) %>% 
  mutate(year = as.numeric(year))
```

# Combined dataset (MMR, GDP per capita, %Healthcare spending, %birth attended by staffs)
```{r}
mmr_gdp_hspend = 
  full_join(mmr_gdp_capita_log, health_spending_percent_gdp, by = c("year", "country", "code"))

mmr_combined = 
  full_join(mmr_gdp_hspend, staff, by = c("year", "country"))

summary(is.na(mmr_combined))
```