---
title: "healthcare_spending_data_cleaning"
author: "Michael Yan"
date: "11/16/2019"
output: html_document
editor_options: 
  chunk_output_type: console
---
# Healthcare spending (%GDP) vs. Country

```{r setup, include = FALSE}
library(tidyverse)
library(rvest)
library(plotly)
library(flexdashboard)
knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

## Import and clean data
```{r}
# Percent GDP that comes from health spending
health_spending_percent_gdp = read_csv("./data/WorldBank/health_spending_percent_GDP.csv") %>%
  janitor::clean_names() %>% 
  select(country_name, country_code, x2000:x2015) %>% 
  pivot_longer(
  x2000:x2015,
  names_to = "year",
  values_to = "health_spending_percent_GDP") %>% 
  rename("health_spending_percent_gdp" = "health_spending_percent_GDP") %>% 
  mutate(year = str_remove(year,"x")) %>% 
  mutate(health_spending_percent_gdp = round(health_spending_percent_gdp, digits = 2)) %>% 
  mutate(country_name = as.factor(country_name),
         country_code = as.factor(country_code),
         year = as.numeric(year)) %>% 
  rename(country = country_name,
         code = country_code)
```

## Filtered dataset to only include 10 developed countries in 2013
```{r}
healthcare_spending_2013_developed = 
  health_spending_percent_gdp %>% 
  filter(year == 2013) %>% 
  filter(country %in% c("United States", 
                        "Switzerland", 
                        "Germany",
                        "Singapore",
                        "Japan",
                        "Canada",
                        "United Kingdom",
                        "Spain",
                        "France", 
                        "Australia"
                        ))
```

## Bar chart (% GDP that comes from healthcare spending in 10 developed countries)
```{r}
ten_developed_countries_health_spending = 
  healthcare_spending_2013_developed %>% 
  mutate(country = as.factor(country)) %>% 
  ggplot(aes(x = country, y = health_spending_percent_gdp, fill = country)) +
  geom_col(position = "dodge") +
  coord_flip() +
  theme(legend.position = "none") +
  labs(x = "Developed countries",
       y = "Percent GDP that comes from healthcare spending(%)",
       title = "Different amount of healthcare spending across 10 developed countries")

interactive_healthcare_spending = ggplotly(ten_developed_countries_health_spending,
                                           tooltip = c("y", "x", "colour"))
```

# Maternal mortality across 10 developed countries

## Import and clean data
```{r}
# Maternal Mortality ratio 
maternal_mort = read_csv("./data/Gapminder/maternal_mortality_ratio_per_100000_live_births.csv") %>% 
  janitor::clean_names() %>% 
  pivot_longer(
    x1800:x2013,
    names_to = "year",
    values_to = "percent_maternal_mortality"
  ) %>% 
  mutate(year = str_remove(year,"x"))
```

## Filtered dataset to only include 10 developed countries in 2013
```{r}
maternal_mort_2013_developed = 
  maternal_mort %>% 
  filter(year == 2013) %>% 
  filter(country %in% c("United States", 
                        "Switzerland", 
                        "Germany",
                        "Singapore",
                        "Japan",
                        "Canada",
                        "United Kingdom",
                        "Spain",
                        "France", 
                        "Australia"
                        ))
```

## Bar chart (maternal mortality in 10 developed countries)
```{r}
ten_developed_countries_maternal_mort = 
  maternal_mort_2013_developed %>% 
  mutate(country = as.factor(country)) %>% 
  ggplot(aes(x = country, y = percent_maternal_mortality, fill = country)) +
  geom_col(position = "dodge") +
  coord_flip() +
  theme(legend.position = "none") +
  labs(x = "Developed countries",
       y = "Maternal mortality(%)",
       title = "Mismatching between US healthcare spending and maternal mortality")

interactive_maternal_mortality = ggplotly(ten_developed_countries_maternal_mort, 
                                          tooltip = c("x", "y"))
```


# Cost of giving birth in 10 developed countries

## Import and clean data
```{r}
# cost of giving birth
child_birth_cost = read_csv("./data/cost_of_child_birth_global.csv") %>% 
  janitor::clean_names() %>% 
  mutate(average_cost = (child_birth_cost_normal + child_birth_cost_csec)/2)
```

## Bar chart (cost of childbirth in 10 developed countires)
```{r}
ten_developed_countries_childbirth_cost = 
  child_birth_cost %>% 
  mutate(country = as.factor(country)) %>% 
  ggplot(aes(x = country, y = average_cost, fill = country)) +
  geom_col(position = "dodge") +
  coord_flip() +
  theme(legend.position = "none") +
  labs(x = "Developed countries",
       y = "Average cost of giving birth(dollars)",
       title = "Cost of child birth in 10 developed countries")

interactive_child_birth_cost = ggplotly(ten_developed_countries_childbirth_cost, 
                                          tooltip = c("x", "y"))
```


## Combine three interactive graphs
```{r}
subplot(interactive_healthcare_spending, 
        interactive_maternal_mortality, 
        shareX = T, shareY = T)

interactive_child_birth_cost

```


