---
title: "Report"
output: 
 html_document:
   theme: flatly
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
library(tidyverse)
library(rvest)
library(plotly)
knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

## Introduction

The United States has one of the highest maternal mortality rate out of developed countries despite high levels of spending. What are the underlying reasons for this? According to the CDC, Maternal Mortality Ratio is defined as the number of pregnancy-related deaths per 100,000 live births in a certain period of time. The CDC reports that in 2016 alone, the US had 16.8 deaths per 100,000 live births compared to 8.6 in the UK. The UK seems to have achieved these results while spending half than what the US spends in healthcare. As a team, we are interested in assessing the maternal mortality ratio against different determining factors to try to understand why the US lags behind other developed nations on this front.


## Data

The data for this project was collected from several sources:

### US Data

The expected cost to the mother of giving birth was scrapped from the following [Business Insider Article](https://www.businessinsider.com/how-much-does-it-cost-to-have-a-baby-2018-4#50-alabama-1). All data for the birth rates and cesearan rates in the US was pulled from the Center for Disease Control and Prevention and can be found at [CDC WONDER](https://wonder.cdc.gov/natality.html). Maternal mortality ratio within the US was derived using the number of maternal deaths, which is also available at from an underlying cause of death dataset at [CDC WONDER](https://wonder.cdc.gov/ucd-icd10.html).


### Global Data

The percent of births attended by a skilled physician was downloaded directly from [Gap Minder](https://www.gapminder.org/data/) which is a collection of datasets from several sources including the World Health Organization. GDP per capita was collected from [Gap Minder](https://www.gapminder.org/data/documentation/). Global healthcare spending or expenditures were collected from [The World Bank](https://data.worldbank.org/indicator/SH.XPD.CHEX.GD.ZS?end=2016&start=2000&view=chart). Cost to the mother was extrapolated from the [International Federation of Health Plans](http://www.ifhp.com/).


## Methods

### Maternal Mortality
The World Health Organization defines Maternal Mortality Ratio (MMR) as the number of maternal deaths per 100,000 births. Since the ratio of maternal death to birth is relatively small, this scaling factor helps to make the data more readable for a human viewer. The global data for MMR was downloaded from [Gap Minder](https://www.gapminder.org/data/).

To clean the global MMR data for use, the data was pivoted from wide to long format. It is also worthwhile to note that this dataset contains data dating back to the 1800s and therefore has many missing values. This is not a problem for our research intentions since we are not fitting regression models for the scope of this problem and therefore we chose to exclude missing values where necessary. Here we are focusing on the maternal moretality in 2013 for 10 randomly selected countries.

```{r MMR_global, results=FALSE, message=FALSE}
#Maternal Mortality ratio from gapminder dataset
maternal_mort = read_csv("./data/Gapminder/maternal_mortality_ratio_per_100000_live_births.csv") %>% 
  janitor::clean_names() %>% 
  pivot_longer(
    x1800:x2013,
    names_to = "year",
    values_to = "percent_maternal_mortality"
  ) %>% 
  mutate(year = str_remove(year,"x"))

maternal_mort_2013_developed = 
  maternal_mort %>% 
  filter(year == 2013) %>% 
  filter(country %in% c("United States", 
                        "Switzerland", 
                        "Germany",
                        "Singapore",
                        "Japan",
                        "Canada",
                        "United Kingdom",
                        "Spain",
                        "France", 
                        "Australia"
                        ))
```

### GDP per capita

GDP per capita stands for Gross Domestic Product per person. It is an important indicator of economic performance and a useful unit to make cross-country comparisons of avereage living standards and economic wellbeing. In our project, we focus on its relationship with Maternal Mortality Ratio (MMR) across countries.

```{r mmr_gdp_per_capita, message = FALSE}
# Maternal Mortality ratio vs. GDP per capita
mmr_gdp_capita = read_csv("./data/maternal-mortality-ratio-vs-gdp-per-capita.csv") %>%
  janitor::clean_names() %>% 
  rename(
    country = entity,
    mmr = maternal_mortality_ratio_deaths_per_100_000_live_births,
    gdp_per_capita = gdp_per_capita_international,
    gdp = x6) %>% 
  mutate(
    country = as.factor(country),
    code = as.factor(code)
  ) %>% 
  mutate(
    gdp_per_capita = round(gdp_per_capita, digits = 2)
  ) %>% 
  filter(year > 1999, year < 2016)
```

While cleaning the data, we renamed variables into a clear format and filtered the dataset from year 2000 to year 2015 to prevent having too many NAs.

```{r ln_gdp_per_capita, message = FALSE}
# log-transformed 'gdp_per_capita'
mmr_gdp_capita_log = 
  mmr_gdp_capita %>% 
  mutate(
    ln_gdp_per_capita = round(log(gdp_per_capita), digits = 2)
  )
```

Particularly, we created a new variable `ln_gdp_per_capita`, the log-transformed `gdp_per_capita`, because GDP per capita has weakness in not taking into account income distribution of a country, and cross-country comparisons based on US dollars can be distorted by exchange rate fluctuations. Therefore, `gdp_per_capita` may not be the best measure of purchasing power, and `ln_gdp_per_capita` was used to reduce the effect of variations. 

### Percent Births Attended By Skilled Health Staff

The percent of births attended by skilled health staff is more explicitly defined as the percentage of deliveries attended by personnel trained to give the necessary supervision, care, and treatment to women during pregnancy, labor, and the postpatrum period. The percent births attended by skilled health staff was taken directly from the aforementioned [Gap Minder](https://www.gapminder.org/data/) dataset. Noteably, this dataset has fewer missing vallues than the MMR dataset, but once again missing values were excluded when necessary.  No complex data manipulation was needed and the following code was used to read in and clean the data:


```{r Births_Attended_SkilledHealthStaff, message=FALSE}
#births attended by skilled health staff (percent of total)
staff = read_csv("./data/Gapminder/births_attended_by_skilled_health_staff_percent_of_total.csv")  %>% 
  janitor::clean_names() %>% 
  pivot_longer(
    x1984:x2017,
    names_to = "year",
    values_to = "perc_births_attended"
  ) %>% 
  mutate(year = str_remove(year,"x"))
```


Both the maternal mortality ratio and births attended by skilled personnel datasets were combined using inner_join.

```{r combine_MMR_attended, message=FALSE}
mortality_births_attend = inner_join(maternal_mort, staff, by = c("country","year"))
```

Since percent births attended by a skilled personnel is a continuous variable, we built a factor `per_range` variable that takes values from 0-100 percent and divide them into four discrete bins. 0-25, 26-50, 51-75, and 76-100 respectively. This new factor variable is then used to plot an interactive boxplot of maternal mortality ratio versus the bins of percent births attended by skilled personnel. To examine whether there was a cutoff past which the percent of births attended by a skilled health physician was no longer a significant factor, we also took a closer look at the countries in this high 76-100% attendance bin.


```{r bin_0to100percent, message=FALSE}
mortality_births_attend = 
mortality_births_attend %>% 
  drop_na() %>% 
  mutate(
    per_range = case_when(
      perc_births_attended <= 25 ~ "0-25",
      perc_births_attended <= 50 ~ "26-50",
      perc_births_attended <= 75 ~ "51-75",
      perc_births_attended <= 100 ~ "76-100",
      TRUE                        ~ ""),
    per_range = factor(per_range, levels = c("0-25", "26-50", 
                                             "51-75", "76-100")),
    text_label = str_c("country: ", country))
```


To do this, we took the `76-100` range and divided it into five smaller ranges and built the same interactive box plot.


```{r bin_top25percent, message=FALSE}
Top_count =
  mortality_births_attend %>% 
  filter(per_range == "76-100") %>%
  mutate(
    top_range = case_when(
      perc_births_attended <= 81 ~ "76-81",
      perc_births_attended <= 85 ~ "82-85",
      perc_births_attended <= 90 ~ "86-90",
      perc_births_attended <= 95 ~ "91-95",
      perc_births_attended <= 100 ~ "96-100",
      TRUE                        ~ ""),
    per_range = factor(per_range, levels = c("76-81", "82-85", 
                                             "86-90", "91-95", "96-100")),
    text_label = str_c("country: ", country)) %>% 
  select(-per_range)
```

### US Cost of Giving Birth

The cost accrued to the mother for giving birth was not readily available in a downloadable dataset. Therefore, we found an article and performed webscraping to extract the data into separate datasets for state and cost for vaginal and cesarean births, with and without insurance:

```{r webscrape_cost_data}
url = "https://www.businessinsider.com/how-much-does-it-cost-to-have-a-baby-2018-4#35-maine-16"

birth_cost_us = read_html(url)

#pull datasets, clean after
state_vec =
  birth_cost_us %>% 
  html_nodes(".slide-title-text") %>% 
  html_text()

all_costs = 
  birth_cost_us %>% 
  html_nodes(".clearfix p+ p") %>% 
  html_text()
```

First, we cleaned the cost data by creating a `cost_df` dataframe using `tibble()` and `janitor::clean_names()`. Next, we further cleaned this dataset as we renamed it `cost_df_clean` by separating the columns using simple expressions and string patterns. After all columns were separated, we recoded the variables and selected variables so that we had a `type` variable containing factors "Vaginal_birth" and "C-section", a `with_ins` variable containing cost with insurance, and a `without_ins` variable containing cost without insurance. Then, we pivoted to long format to create a variable `insurance` with factors `with_ins` or "with insurance" and `without_ins` or "without insurance" and `cost` which details the cost in US dollars. At this point, it is understood that there are 4 unique entries per state, although state name is not yet included in the dataset.

```{r clean_US_cost_data}
#pivot longer so all costs are in one column
cost_df = tibble(
  all_costs
) %>% 
  janitor::clean_names()

cost_df_clean = cost_df %>% 
    separate(all_costs, into = c("delivery_method", "other", "cost_wo"), sep = "\\:") %>% 
    separate(other, into = c("cost", "empty"), sep = "C-section without insurance" ) %>% 
    separate(cost, into = c("cost_w", "empty"), sep = "Vaginal birth without insurance")%>%
  select(-empty) %>% 
  mutate(
    type = factor(delivery_method), 
    type = recode(type,
    "Vaginal birth with insurance" = "Vaginal_birth",
    "C-section with insurance" = "C-section")
  ) %>% 
  select(-delivery_method) %>% 
  rename("with_ins" = "cost_w", #fix mislabeled data
             "without_ins" = "cost_wo") %>% 
  select(type,with_ins,without_ins) %>% 
  pivot_longer(
    with_ins: without_ins,
    names_to = "insurance",
    values_to = "cost"
  )
  
```


Next, we used `tibble()` to create a dataframe, in which we cleaned the vector state names, preserving their import order. We got rid of the string which numbered the states. Then, we created variables for `Vaginal_birth`, `C-section`, `with_ins` and `without_ins`. Next, we used pivoted to long format with `Vaginal` and `C-section` as factors of the variable `type` and `with_ins` and `without_ins` as factors of the variable `insurance`. This created a dataset with each state having 4 entries for vaginal and cesarean births, both with and without insurance. Then, we deselect the two dummy variables created, which were filled with placeholder numbers. This decision was made to create a dataset with the same format as the `cost_df`. 

```{r clean_state_names}
#clean state names
state_df = tibble(
  state = state_vec %>% 
    str_replace(".\\. ","") %>%
    str_replace("[1-9]",""),
  "Vaginal_birth" = 1,
  "C-section" = 0,
  "with_ins" = 2,
  "without_ins" = 3
) %>% 
  pivot_longer(
    "Vaginal_birth" : "C-section",
    names_to = "type",
    values_to = "dummy"
  ) %>% 
  pivot_longer(
    "with_ins" : "without_ins",
    names_to = "insurance",
    values_to = "dummy2"
  ) %>% 
  select(state, type, insurance, -dummy, -dummy2)
```

Finally, we merged these two datasets. Unfortunately, web scraping gets a bit messy. We had to assume the order of the entries was preserved, which was verified for the first few entries. Therefore, we simply mutated to create a new variable and pasted the state variable from the state dataset into the first column of the cost dataset to complete our `US_cost_to_mother_df` dataframe. This is the final clean dataset which was saved as an excel file to more easily access results for the rest of this project.

```{r merge_country_cost_savecsv}
# merge the state information with the cost information
# write to csv

US_cost_to_mother_df = 
  cost_df_clean %>% 
  mutate(state = state_df$state) %>% 
  select(state,type, insurance, cost)

write.csv(US_cost_to_mother_df, file = "US_cost_to_mother_df.csv")

```



### US Cesarean Rates
Since the cesarean rates were not directly available from the CDC, we first downloaded the number of births and filtered vaginal and cesearean deliveries in each state. By taking the number of cesarean deliveries and dividing by the total number of births per state, we extrapolated the cesaraean rate for each state. 


First, we read in the data and performed basic cleaning using `janitor::clean_names()`. Then we filtered by birth delivery method (`delivery_method`), omitting those which were unknown. The `births` were automatically imported as factors so the `births` had many errors when they were converted directly to numeric. To solve this, we converted the factors to characters and then to numeric variable types.

```{r births_per_delivery_method_per_state}
#load number of births per delivery method per state
birth_type =
  read.delim("./dataset_births_us/birth_delivery_by_state.txt") %>% 
  janitor::clean_names() %>% 
  filter(delivery_method != "Unknown or Not Stated") %>% 
  na.omit() %>% 
  mutate(births = as.numeric(as.character(births))) %>% 
  select(state_of_residence, delivery_method, births) 
```


Then, we calculated the total number of births per state by using group and summarize.

```{r total_births_per_state}
#calculate total number births per state
birth_sum_by_state =
  birth_type %>%
  group_by(state_of_residence) %>% 
  summarize(
    total_births = sum(births)
    ) %>% 
  ungroup()
```


Finally, we joined these two datasets using a `left_join` by state of residence. Then, we used the number of cesarean births divided by the total number of births per state to produce a cesarean rate per state.

```{r Cesarean_rates_per_state}
#calculate percent of c-sections per state by 1st merging and then creating a new variable
birth_df =
  left_join(birth_type, birth_sum_by_state, by = "state_of_residence") %>% 
  filter(delivery_method == "Cesarean") %>%
  select(-delivery_method) %>% 
  mutate(c_rate  = births/total_births) %>% 
  select(state_of_residence, total_births, c_rate)
```


### US State Maternal Mortality Ratio

The Maternal Mortality ratios in the US were not directly available. To circumvent this, we used the number of maternal deaths per state and divided this by the number of births per state to get a ratio of maternal deaths per birth. Then, we scaled this number by multiplying by 100,000 which gives the number of maternal death per 100,000 births for consistency with the statistics reported by the World Health Organization.

To arrive at this number, we first cleaned the underlying cause of death dataset:

```{r}
cdc_maternal_death_data = read.delim("./us_shiny/data/maternal_deaths_1999_2017.txt") %>%
  janitor::clean_names() %>%
  na.omit() %>%
  subset(notes != "Total") %>%
  select(year, state, deaths)

```



Next, we cleaned the total number of births per state from the natality dataset:

```{r}
births_by_state =
  read.delim("./dataset_births_us/Natality_2016-2018_expanded.txt") %>% 
  janitor::clean_names() %>% 
  select(state_of_residence,births,total_population)
```

These numbers are off ever so slightly from the data used to calculate cesarean rates. For example the births in alabama differ by `r pull(filter(births_by_state,state_of_residence=="Alabama"),births) - pull(filter(birth_df,state_of_residence=="Alabama"),total_births)` people. This difference may be attributed to the data which was supressed because the delivery method of the birth was not known. This is expected to be because we excluded the "unknown" delivery method in the natality dataset to more accurately calculate the percent of births delivered via cesarean section to the best of our knowledge with the dataset provided. To calculate maternal mortality ratio, we chose to use the total number of births including those with an unknown delivery method, since it makes sense to include this metric when calculating a mortality ratio where we are not specifically considering delivery method.

It is also important to note that the number of births was only calculated from 2018 whereas the underlying cause of death dataset extends from 1999-2017. Given the time constraints of this project, the number of births per state for 2018 was used to normalize the deaths for every year from 1999-2017. However, we acknowledge that it would be better practice to calculate the number of births in each year from 1999-2017 and use this year-specific number when calculating the maternal mortality ratio (per 100,000 births) in each of these years.


### Percent GDP that comes from healthcare spending
Global healthcare spending or expenditures were collected from [The World Bank](https://data.worldbank.org/indicator/SH.XPD.CHEX.GD.ZS?end=2016&start=2000&view=chart). This shows the healthcare spending's contribution to the total GDP. The dataset is relatively clean. What we did was renaming variables, filtering out the 10 selected countries in 2013 and pivoting from wide to long format. In our project, we are trying to see if percent GDP that comes from healthcare spending is an indicator for maternal mortality. We chose 2013 because it is the most recent year that we have both the percent GDP that comes from healthcare spending data and the maternal mortality data.

```{r healthcare spending percent GDP, message=FALSE}
health_spending_percent_gdp = read_csv("./data/WorldBank/health_spending_percent_GDP.csv") %>%
  janitor::clean_names() %>% 
  select(country_name, country_code, x2000:x2015) %>% 
  pivot_longer(
  x2000:x2015,
  names_to = "year",
  values_to = "health_spending_percent_GDP") %>% 
  rename("health_spending_percent_gdp" = "health_spending_percent_GDP") %>% 
  mutate(year = str_remove(year,"x")) %>% 
  mutate(health_spending_percent_gdp = round(health_spending_percent_gdp, digits = 2)) %>% 
  mutate(country_name = as.factor(country_name),
         country_code = as.factor(country_code),
         year = as.numeric(year)) %>% 
  rename(country = country_name,
         code = country_code)

healthcare_spending_2013_developed = 
  health_spending_percent_gdp %>% 
  filter(year == 2013) %>% 
  filter(country %in% c("United States", 
                        "Switzerland", 
                        "Germany",
                        "Singapore",
                        "Japan",
                        "Canada",
                        "United Kingdom",
                        "Spain",
                        "France", 
                        "Australia"
                        ))
```

### Cost of giving birth in 10 developed countries
Cost of giving birth was extrapolated from the [International Federation of Health Plans](http://www.ifhp.com/). The dataset is clean and we didn't do much of cleaning. We did create a new variable showing the average cost of getting a normal delivery and getting a C-section since the costs are different. What we are trying to see here is whether higher spending on a family scale leads to lower maternal mortality. 

```{r cost of giving birth, message=FALSE}
child_birth_cost = read_csv("./data/cost_of_child_birth_global.csv") %>% 
  janitor::clean_names() %>% 
  mutate(average_cost = (child_birth_cost_normal + child_birth_cost_csec)/2)
```


## Results

```{r scatterplot_mmr_gdp}
x1_template = list(
  title = "GDP per capita ($)"
)

x2_template = list(
  title = "ln(GDP per capita)"
)

y_template = list(
  title = "Maternal Mortality Ratio per 100,000 live births"
)

mmr_gdp_capita_log %>%  
  filter(year == 2013) %>% 
  mutate(text_label = str_c("GDP per capita ($): ", gdp_per_capita)) %>% 
  plot_ly(
    x = ~ln_gdp_per_capita, y = ~mmr, type = "scatter", 
    mode = "markers", color = ~country, text = ~text_label
  ) %>%  
  layout(
    xaxis = x2_template,
    yaxis = y_template,
    title = "Maternal mortality ratio vs. ln(GDP per capita)",
    showlegend = FALSE
  )
```

The scatterplot shows that globally in 2013, `ln(GDP per capita)` is approximately inversely associated with `Maternal Mortality Ratio per 100,000 live births`. From this overview, we see that GDP per capita can be a strong predictor of maternal mortality ratio at global level.

### Maternal Mortality in 10 Developed Countries

```{r plot mortality global, message=FALSE}
ten_developed_countries_maternal_mort = 
  maternal_mort_2013_developed %>% 
  mutate(country = as.factor(country),
         country = fct_reorder(country, percent_maternal_mortality)) %>% 
  ggplot(aes(x = country, y = percent_maternal_mortality, fill = country)) +
  geom_col(position = "dodge") +
  coord_flip() +
  theme(legend.position = "none") +
  labs(x = "Developed countries",
       y = "Maternal Mortality Ratio (deaths per 100,000 births)"
       )

interactive_maternal_mortality = ggplotly(ten_developed_countries_maternal_mort, 
                                          tooltip = c("x", "y"))
interactive_maternal_mortality
```

Now, we focus specifically on the top 10 developed countries. As observed, the US has the highest maternal mortality ratio among the ten selected countries. 

```{r}
ten_developed_countries_gdp_per_capita = 
mmr_gdp_capita_log %>%  
  filter(year == 2013,
         country %in% c("United States", 
                        "Switzerland", 
                        "Germany",
                        "Singapore",
                        "Japan",
                        "Canada",
                        "United Kingdom",
                        "Spain",
                        "France", 
                        "Australia"
                        )) %>% 
  mutate(country = as.factor(country),
         country = fct_relevel(country, 
                               "Spain",
                               "Australia",
                               "Japan",
                               "Singapore",
                               "Switzerland",
                               "Germany",
                               "United Kingdom",
                               "Canada",
                               "France",
                               "United States")) %>% 
  ggplot(aes(x = country, y = gdp_per_capita, fill = country)) +
  geom_col(position = "dodge") +
  coord_flip() +
  theme(legend.position = "none") +
  labs(x = "Developed countries",
       y = "GDP per capita (dollars)",
       title = "GDP per capita in top 10 developed countries")

interactive_gdp_per_capita = ggplotly(ten_developed_countries_gdp_per_capita, 
                                          tooltip = c("x", "y"))
interactive_gdp_per_capita
```

Together with the maternal mortality ratio plot, we see a general trend that the higher the GDP per capita, the lower the maternal mortality ratio. However, countries like United States, France and Canada, have both high maternal mortality ratio and high GDP per capita. This indicates that the economic performance and average living standards of an individual or a country may not thoroughly reflect how well its maternal health system functions.

```{r plot healthcare spending, message=FALSE}
ten_developed_countries_health_spending = 
  healthcare_spending_2013_developed %>% 
  mutate(country = as.factor(country),
         country = fct_relevel(country, 
                               "Spain",
                               "Australia",
                               "Japan",
                               "Singapore",
                               "Switzerland",
                               "Germany",
                               "United Kingdom",
                               "Canada",
                               "France",
                               "United States")) %>% 
  ggplot(aes(x = country, y = health_spending_percent_gdp, fill = country)) +
  geom_col(position = "dodge") +
  coord_flip() +
  theme(legend.position = "none") +
  labs(x = "Developed countries",
       y = "Percent GDP that comes from healthcare spending (%)"
       )

interactive_healthcare_spending = ggplotly(ten_developed_countries_health_spending,
                                           tooltip = c("y", "x", "colour"))
interactive_healthcare_spending
```

Based on the percent GDP plot and the previous maternal mortality plot, despite spending the most amount of money, the US has the highest maternal mortality ratio. There is an obvious mismatch between the amount of money spending on healthcare and the maternal mortality which indicates that there are problems waiting to besolved in the healthcare system.

```{r cost of giving birth in 10 developed countries, message=FALSE}
ten_developed_countries_childbirth_cost = 
  child_birth_cost %>% 
  mutate(country = as.factor(country),
         country = fct_relevel(country, 
                               "Spain",
                               "Australia",
                               "Japan",
                               "Singapore",
                               "Switzerland",
                               "Germany",
                               "United Kingdom",
                               "Canada",
                               "France",
                               "United States")) %>% 
  ggplot(aes(x = country, y = average_cost, fill = country)) +
  geom_col(position = "dodge") +
  coord_flip() +
  theme(legend.position = "none") +
  labs(x = "Developed countries",
       y = "Average cost of giving birth ($)"
       )

interactive_child_birth_cost = ggplotly(ten_developed_countries_childbirth_cost, 
                                          tooltip = c("x", "y"))
interactive_child_birth_cost
```

Based on the plot, there is not an obvious relationship between cost of giving birth and maternal mortality.



### Birth Attended by skilled Staff



```{r plot_attended, message=FALSE}
mortality_births_attend %>% 
  plot_ly(y = ~percent_maternal_mortality, color = ~per_range, type = "box",
          colors = "Set2", text = ~text_label) %>% 
  layout(
    xaxis = list(title = "Birth Attended by Skilled Staff (% Range)"),
    yaxis = list(title = "Maternal Mortality per 100k Live Births"))
```


As expected, the maternal mortality decreases as births attended by skilled staff increases. Now we will direct our attention to countries with higher percentage of skilled staff attending births. 

```{r, plot_attended_top25percent, message=FALSE}
Top_count %>%
  plot_ly(y = ~percent_maternal_mortality, color = ~top_range, type = "box",
          colors = "Set2", text = ~text_label) %>% 
  layout(
    xaxis = list(title = "Birth Attended by Skilled Staff (%)"),
    yaxis = list(title = "Maternal Mortality per 100k Live Births"))
```

There does not appear to be a significant trend but there is an overall decrease in maternal mortality ratio as the births attendance increases.



## Discussion

Maternal mortality is a complex health issue impacting societies around the world. In this research study, we analyzed some of many factors determining pregnancy outcomes in different countries. Factors such as health care spending, GDP per capita, the percentage of births attended by a nurse or physician, and how much it costs a mother to give birth.

Overall, we found that maternal mortality ratio decreases as births attended by skilled staff or GDP per capita increases. But interestingly enough, we found that for the wealthiest nations, maternal mortality was not determined by a country's GDP per capita. For instance, France was found to have a lower GDP per capita compared to the US but also had a far lower maternal mortality ratio (12 deaths per 100,000 live births) compared to the US (28 deaths per 100,000 live births). This was an unexpected observation considering the global relationship seen between GDP per capita and maternal mortality.

Another observation worth highlighting is that even though healthcare makes up a large portion of the GDP in the US, this country still has the highest maternal mortality ratio when compared to other developed nations. In the future, it may be more meaningful to investigate the amount of government spending the US invests in its healtchare system rather than GDP.

When considering the cost to the family, it may be a more reasonable metric to normalize the cost of giving birth by the average salary in each location. This would give a more fair representation of the cost burden around the world.

It would also be interesting to look at differing maternal mortality ratios between populations, those who have cesarean sections and those who do not. However, the population specific data to perform these studies is not available at this time.

Given that many complications arise before and after birth, prenatal care and family leave at the time of childbirth may also play a significant role in maternal outcomes. 

While this project serves as a basis of understanding, more work is urgently needed to complete our understanding of the factors driving maternal mortality. This is needed to better understand why the US lags behind other first world countries and explain deviations in maternal mortality outcomes worldwide.
